Beyesian Inference for Decision Making under COVID-19: A case study

### 2020-05-31 out/paper.Rmd
corrección ortografía y draft seccion VII

---
title: "Beyesian Inference for Decision Making under COVID-19: A case study"
subtitle: ""
author: ""
header-includes: 
  - \usepackage{fancyhdr}
date: ""
output: 
   pdf_document:
    toc: FALSE
    highlight: 'kate'
    number_sections: TRUE
editor_options: 
  chunk_output_type: console
abstract: "Inference on circumstances where hypothesis have to be done with no data to substantiate them is tough. The following paper presents a case study on COVID-19 induced job loss and how the bayesian theory of inference sets clear ground rules for incorporating such hypothesis and the caveats of doing so with the Mathematical Statistics approach"
---

\pagenumbering{arabic} 
\fancyhead[L]{\thepage}
\fancyhead[C]{}
\fancyhead[R]{}
\fancyfoot[L]{Haro}
\fancyfoot[R]{Peniche}
\pagestyle{fancy}
\addtolength{\headheight}{1.0cm}
\pagestyle{fancyplain}

@@

The worldwide health crisis induced by the the ' *SARS-COV 12* ', globally referred as COVID-19, mutated the classic decision making under uncertainty problem into a substancially more uncertain mess due to the lack of closeness of the 2020 pandemic to previous world plights. Under this particuarly uncertain context Bayesian Theory of Inference bestows an adecuate framework to incorporate such new precariousness.

In particular, having a reliable ground where to stand becomes critical for policy makers or other high impact decision makers. The following paper presents a simple, yet non trivial, case study of Bayesian Inference to adress the difficulty of predicting the job loss induced by the COVID-19 lockdown measures. We also stress out some of the caveats of the classical aproach from the Fisher and Neyman school 
of Mathematical statistics and contrast both approaches slanting outyurns in the context of desicion making.

Section **I** begins by providing the reader a concrete outlook of the problem and the data at hando to tackle it. Section **II** translates the problem into its concrete mathematical and statistical representation and comments on the frailty of classical solution. Section **III** goes about the same formulation under a simple *by-hand* bayesian alternative and introduces the concept ok likelihood resilience, while section **IV** proposes a simpler solution using an alternative conjugate family. Section **V** presents a more sophisticated approach through Monte Carlo Markoc Chain Stochastic simulation an finally section **VI** is a brief utterance on the importance of providing a well grounded inference for decision making proceses.

@@

Let $N-t$ be the number of workers noticed on time or in advanced by some company to the Social Service Mexican Insitute at some given month $t$. As mentioned before, it's of interest to predict the amounts of the labor force $N_{t+1}$ for the succesive month at the instances of the health crisis.

Which could be a sufficiently simple formulation of the problem? At first, the notation used and even the timely nature of the phenom may suggest some time series approach. However it´s the objective of the authors to present a as simple as posible formulation for the model so. Being that so, though it may look abit naive at first, lets consider the following structural form for the lagged worker numbers at $t+1$  being: $$N_{t+1} = N_{t} \cdot ( 1 + \theta)$$

@@

Yet even more naive and for the sake of simplicity lets assume $x_i = (\frac{N_{t+1}}{N_t} - 1) \sim N(\theta, \sigma^2)$ and that $\{x_i \}_{i=1}^N$ constitutes a colection of $i.i.d$ random variables i.e. a *random sample* $X_{(\underline{n})}$. Proceding after the construction of this model $\hat{\theta}$ and $\hat{\sigma}^2$ would correspond to the MLE estimators nothing new.

Although the normal parametrization is convenient for its simplicity, it should stand out to anyone that there is a major problem choosing the gaussian family because $\theta$ $\epsilon$ $\Theta = [ 0, \infty ]$ and under the normality assumption $\Theta =  [- \infty, \infty]$. For now, our way around will be to truncate such $N(\hat{\theta}, \hat{\sigma^2})$ and redistribute tail density [^1].

@@

In any context not involving such distressfull times everything would work fine, in fact our Fisherian conclusion would be that with 95% confidence level $\theta$ would lie within the **INSERTAR INTERVALO** [^2]. However things start getting funny when incorporating unobserved hypothesis on the probable behaiviour of $\theta_{COVID-19}$. Lets explore some alternatives:

[^2]: This interval lenght corresponds to the *predictive interval lenght* given by $\frac{X_{n+1} - \overline{X}}{s_n \sqrt{1 + \frac{1}{n}}} \sim T^{n-1}$. The interval is not symetric because of the truncation.

  a. Let Marcos be some well trained senior economist Ph.D on labor market dynamics, well known for his work as a government advisor for the Mexican Ministry of Labor. As an experienced fellow and given the experience in similar (or not that similar countries) Marcos believes, in fact he is almost sure, that $\theta_{COVID-19}$ will in fact be of around 5%. Still, the only way to reach 5% levels given the MLE framework would be arbitrary marking down *a posteriori* $\hat{\theta}_{COVID-19}$ under the pretence of having strong evidence in favour of 5%.

Many coleagues would argue that this textbook like toy model doesn´t even correspond to ordered nature of observations or even the parametral space. However any other, more sophisticated, model would encounter the same incovenience as there would be no input accounting for COVID-19 induced contraction at the time, not to mention the rise in model complexity and the need for computer power.

Although Marcos' opinion may be well founded and it could be right, it appears to be introduced into the inference machinery by coercing the likelihood giving the prior not voice nor vote. Additionaly this *mark down* fashion of placing constrains on the data is not accompanied by the corresponding increase in standard deviation due to the rise in uncertainty of the actual $\theta_{COVID-19}$ behaviour.

  b. Katia, a fellow colleague of Marcos, suggests increasing $\sigma_{MLE}^2$ by some amount to reflect the uncertainty. Marcos agrees with katia, however the question that will innevitably arise is, by how much?.

Again, any given arbitrary quantity for increasing variance will be an exogenous input in the classical inference workflow. Although there may be certain rules or thumb from which the writers are currently unaware, in the context of decition making environments an inference which could structurally and consistently incorporate both Marcos' and Katia's insight would be preferred over case specific tips.

@@

Rescuing our "not-so-toy" example, an alternative to adress some of the classical framework caveats is to proceed through bayesian grounds. We'll part from the same assumptions and the same structural construction from the model except, for noew, we will asume variance to be known and equal to historical variance, additionally we assume a reference distribution for $\theta$ given by $N(\theta | \mu_0, \sigma_0^2)$ in orider to take advantage of a conjugacy scheme. Building the model this way, we get that the posterior predictive distribution for $\theta_{t+1} = \theta_{COVID-19}$ is given by $$f(X_{t+1}| X_{(\underline{n})}) \sim N(x | \mu_{n}, \sigma_{n}^2 + \sigma_{x}^2)$$

@@

Commenting on the forgoing model, it could de said that we gain complexity at the cost os assuming a known variance, which in obviously not true. Nevertheless we gained so much more more than a little mathematical complexity, this being a consistent way to introduce Marco's expert knowledge into the inferential process as an endogenous input through $\theta$ 's *a priori* distribution.

Following this line of though, Marco's expert knowledge is introduced in the form of a (highly) **informative prior**. Expertise is incorporated first by centering the future bahaviour of the lagged number of ployees around 5% and second by adjusting the level of "surenees" by declaring a very low standard deviation [^3] lets say .001%. In fact chossing this particular characterization for $\theta_{COVID-19}$ 's prior yields an prediction credibility interval of **INSERTAR INTERVALO DE CREDIBILIDAD** around **INSERTAR ESTIMACIÓN PUNTUAL** [^4].

@@

Even though this specific prior characterization is as arbitrary as any number choosen for a *mark-down procedure* the *a prior* and not *a posteriori* fashion of introducing this knowledge, plus the interaction between paramters through the conjugacy scheme and the **likelihood resilience** [^5] allow for a coherent manner to let hypothesis and data interact with each other parsimonious fashion within the walls of a model[^6].

@@

From last paragraph its particuarly worth discussing the term *likelihood resilience* from the known variance Normal-Normal conjugate family. Which is formally measured as the relative in change in the posterior distributions parameters in the presence of extreme prioir **FALTA PULIR ESA DEFINICIÓN CON MEDIDAS RELATIVAS ESPECÍFICAS EN TERMINOS DE MOVIMIENTOS EXTREMOS DE LA PRIOR PARA CIERTO TRHESHOLD**. In Figure 1 we can aprreciate the posteriors distribution behaviour with different prior paramters.

@@

**DESCRIBIR COMPORTAMIENTO**. In other words likelihood resilience in this model is not just an intrinsic characteristic that prevents arbitrary data coercion or the *lying with statistics* issue but a quantifiable measure of how robust the inference is against extreme hypotheis.

Nonetheless many may be still be troubled by the known variance assumption, which is in fact totally reasonable. Relaxing this premise is just a matter of a litter more mathematical intricacy [^7]. For a more realistic scenario we choose the Normal-InverseGamma family where the predictive distribution for $X_{n+1}$ is: $$f(X_{na+1} | X_{(\underline{n})}) \sim  t_{2 \alpha_n}(x | \mu_n, \frac{\beta_n( \kappa_n + 1)}{\alpha_n \kappa_n})$$
Where:

@@

As in the known variance case described some lines earlier, we need to truncate the posterior predictive distribution to adjust for $\theta$ 's parametral space yielding the following 95% credibility interval of **INSERTAR INTERVALO** [^8].

[^8]: Truncation of posterior predictive t-distribution was dode with r package *truncdist* v. 1.0-2.

@@

Although analytical manipulation become a bit more messy than obtainign the MLE from the classical aproach, it should be obvious by know the benefit of advocating for the later methodology. We gain a coherent inference substructure to include unobserved hypotheis, we get the likelihood resilience insurance for free and in the conjugte case we can analytically interaction between model paramters (which in the mark-down fashion we cannot).

@@

We finally need to adress the parametral space issue as truncation may introduce some distortions of the interest paramter's behaviour. To account for a more realistic representation of $\Theta$, in which case we could opt to characterize $X_i$ as $Exp(\theta)$ however the MLE will still have problem **a** and **b** as described in section I.

@@

Given this second conjugate scheme we account for the actual $\Theta$, in fact we gains a substntially more simple math than the normal conjugate case and again we acquiere the likelihood resilience insurance for free. Under this new parametrization and setting $\alpha_0 =$ **INSERTAR NÚMERO** and $\beta_0 =$ **INSERTAR NÚMERO** we obtain a 95% credible interval of **INSERTAR INTERVALO**.

Again we can directly analyse likelihood resilience to different priors yielding figure 3.

@@

>>>>

---
title: "Beyesian Inference for Decision Making under COVID-19: A case study"
subtitle: ""
author: ""
header-includes: 
  - \usepackage{fancyhdr}
date: ""
output: 
   pdf_document:
    toc: FALSE
    highlight: 'kate'
    number_sections: TRUE
editor_options: 
  chunk_output_type: console
abstract: "Inference on circumstances where hypothesis have to be done with no data to substantiate them is taugh. The following paper presents a case study on COVID-19 induced job loss and how the bayesian theory of inference sets clear ground rules for incorporating such hypothesis and the caveats of doing so with the Mathematical Statistics approach"
---

\pagenumbering{arabic} 
\fancyhead[L]{\thepage}
\fancyhead[C]{}
\fancyhead[R]{}
\fancyfoot[L]{Carlos Haro}
\fancyfoot[R]{Gibrán Peniche}
\pagestyle{fancy}
\addtolength{\headheight}{1.0cm}
\pagestyle{fancyplain}

@@

The worldwide health crisis induced by the the " *SARS-COV 12* ", globally referred as COVID-19, mutated the classic decision making under uncertainty problem into a substancially more uncertain mess due to the lack of closeness of the 2020 pandemic to previous world plights. Under this particularly uncertain context Bayesian Theory of Inference bestows an adequate framework to incorporate such new precariousness.

Having a reliable ground where to stand becomes critical for policy makers or other high impact decision makers. The following paper presents a simple, yet non trivial, case study of Bayesian Inference to address the difficulty of predicting the job loss induced by the COVID-19 lock down measures. We also stress out some of the caveats of the classical approach from the Fisher and Neyman school 
of Mathematical statistics and contrast both approaches slanting outurns in the context of decision making.

Section **I** begins by providing the reader a concrete outlook of the problem and the data at hand to tackle it. Section **II** translates the problem into its concrete mathematical and statistical representation and comments on the frailty of classical solution. Section **III** goes about the same formulation under a simple *by-hand* bayesian alternative and introduces the concept of likelihood resilience, while section **IV** proposes a simpler solution using an alternative conjugate family. Section **V** presents a more sophisticated approach through Monte Carlo Markov Chain Stochastic simulation an finally section **VI** is a brief utterance on the importance of providing a well grounded inference for decision making processes.

@@

Let $N_t$ be the number of workers noticed on time or in advanced by some company to the Social Service Mexican Institute at some given month $t$. As mentioned before, it's of interest to predict the amounts of the labor force $N_{t+1}$ for the successive month at the instances of the health crisis.

Which could be a sufficiently simple formulation of the problem? At first, the notation used and even the timely nature of the phenom may suggest some time series approach. However it´s the objective of the authors to present a as simple as possible formulation for the model. Being that so, though it may look a bit naive at first, lets consider the following structural form for the lagged worker numbers at month $t+1$, that being: $$N_{t+1} = N_{t} \cdot ( 1 + \theta)$$

@@

Yet even more naive and for the sake of simplicity lets assume $x_i = (\frac{N_{t+1}}{N_t} - 1) \sim N(\theta, \sigma^2)$ and that $\{x_i \}_{i=1}^N$ constitutes a colection of $i.i.d$ random variables i.e. a *random sample* $X_{(\underline{n})}$. Going further $\hat{\theta}$ and $\hat{\sigma}^2$ would correspond to the MLE estimators, nothing new.

Although the normal parametrization is convenient for its simplicity, it should stand out to anyone that there is a major problem choosing the Gaussian family, this is so because $\theta$ $\epsilon$ $\Theta = [ 0, \infty ]$ and under the normality assumption $\Theta =  [- \infty, \infty]$. For now, our way around will be to truncate such $N(\hat{\theta}, \hat{\sigma^2})$ and redistribute tail density [^1].

@@

In any context not involving such distressful times everything would work fine, in fact our Fisherian conclusion would be that with 95% confidence level $\theta$ would lie within the **INSERTAR INTERVALO** [^2]. However things start getting funny when incorporating unobserved hypothesis on the probable behavior of $\theta_{COVID-19}$. Lets explore some alternatives:

[^2]: This interval length corresponds to the *predictive interval length* given by $\frac{X_{n+1} - \overline{X}}{s_n \sqrt{1 + \frac{1}{n}}} \sim T^{n-1}$. The interval is not symmetric because of the truncation.

  a. Let Marcos be some well trained senior economist Ph.D on labor market dynamics, known for his work as a government adviser for the Mexican Ministry of Labor. As an experienced scientist and given the experience in similar (or not that similar countries) Marcos believes, in fact he is almost sure, that $\theta_{COVID-19}$ will in fact be of around 5%. Still, the only way to reach 5% levels given the MLE framework would be arbitrary marking down *a posteriori* $\hat{\theta}_{COVID-19}$ under the pretence of having strong evidence in favor of 5%.

Many colleagues would argue that this textbook like toy model doesn´t even correspond to the ordered nature of observations or even the parametric space. However any other, more sophisticated, model would encounter the same inconvenience as there would be no input accounting for COVID-19 induced contraction at the time, not to mention the rise in model complexity and the need for computer power.

Although Marcos' opinion may be well founded and it could be right, it appears to be introduced into the inference machinery by coercing the likelihood's output giving it not voice nor vote. Additionally this *mark down* fashion of placing constrains on the data is not accompanied by the corresponding increase in standard deviation due to the rise in uncertainty of the actual $\theta_{COVID-19}$ behavior.

  b. Katia, a fellow colleague of Marcos, suggests increasing $\sigma_{MLE}^2$ by some amount to reflect the uncertainty. Marcos agrees with Katia, however the question that will inevitably arise is, by how much?.

Again, any given arbitrary quantity for increasing variance will be an exogenous input in the classical inference work flow. Although there may be certain rules of thumb from which the writers are currently unaware, in the context of decision making environments an inference which could structurally and consistently incorporate both Marcos' and Katia's insight would be preferred over case specific tips.

@@

Rescuing our "not-so-toy" example, an alternative to address some of the classical framework caveats is to proceed through bayesian methodss. We'll part from the same assumptions and the same structural construction from the model except, for now, we will assume variance to be known and equal to historical variance, additionally we assume a reference distribution for $\theta$ given by $N(\theta | \mu_0, \sigma_0^2)$ in order to take advantage of a conjugate scheme. Building the model this way, we get that the posterior predictive distribution for $\theta_{t+1} = \theta_{COVID-19}$ is given by $$f(X_{t+1}| X_{(\underline{n})}) \sim N(x | \mu_{n}, \sigma_{n}^2 + \sigma_{x}^2)$$

@@

Commenting on the forgoing model, it could de said that we gain complexity at the cost of assuming a known variance, which in the authors' opinion is definitely not true. Nevertheless we gained so much more more than a little mathematical complexity, that being a consistent way to introduce Marcos' expert knowledge into the inferential process as an endogenous input, rather than an exogenous distortion on the model, through $\theta$ 's *a priori* distribution.

Following this line of though, Marcos' expert knowledge is introduced in the form of a (highly) **informative prior**. Expertise is incorporated first by centering the future bahaviour of the lagged number of employees around 5% and second by adjusting the level of "sureness" by declaring a very low standard deviation [^3] lets say .001%. In fact chossing this particular characterization for $\theta_{COVID-19}$ 's prior yields a prediction intervala at 95% credibility level of **INSERTAR INTERVALO DE CREDIBILIDAD** around **INSERTAR ESTIMACIÓN PUNTUAL** [^4].

@@

Even though this specific prior characterization is as arbitrary as any number chosen for a *mark-down procedure* the *a priori* and not *a posteriori* fashion of introducing this knowledge, plus the interaction between parameters through the bayesian inference frame work and the **likelihood resilience** [^5] allow for a coherent manner to let hypothesis and data interact with each other in a parsimonious fashion within the walls of a model[^6].

@@

From last paragraph its particularly worth discussing the term *likelihood resilience*. This formally defined as the **sensitivity of the posterior paramaters to extreme prior distributions** or in other words "how strong is the likehood evidence of the data".  Accounting for likelihood resilience is done by quantifying the relative change in the posterior distributions parameters from the original hypothesis by in the presence of extreme prior **FALTA PULIR ESA DEFINICIÓN CON MEDIDAS RELATIVAS ESPECÍFICAS EN TERMINOS DE MOVIMIENTOS EXTREMOS DE LA PRIOR PARA CIERTO TRHESHOLD**. Likelihood resilience can be though as an end-user type off insurance, in the waythat  reporting this metric along with the results prevents, at least in some degree, the *"lying with statistics phenom"*. Further more its an objective measure of the robustness of the inference done in presence of an unobserved hypothesis. In Figure 1 we can appreciate the posterior's distribution behavior with different prior paramaters.

@@

**DESCRIBIR COMPORTAMIENTO**. Likelihood resilience is an intrinsic characteristic arising from the way our inference was built. additionaly in can be analytically traced and studied in conjugate schemes through the functional form of the estimators and numerically when applying MCMC methods.

Nonetheless many may be still be troubled by the known variance assumption, which is totally reasonable. Relaxing this premise is just a matter of a little more mathematical intricacy [^7]. For a more realistic scenario we choose the Normal-Inverse Gamma family where the predictive distribution for $X_{n+1}$ is: $$f(X_{n+1} | X_{(\underline{n})}) \sim  t_{2 \alpha_n}(x | \mu_n, \frac{\beta_n( \kappa_n + 1)}{\alpha_n \kappa_n})$$
Where:

@@

We set prior hyperparamters to be **Insertar supuestos** to reflect a similar uncertainty as in the first case. As in the known variance case described some lines earlier, we need to truncate the posterior predictive distribution to adjust for $\theta$ 's parametric space yielding the following 95% credibility interval of **INSERTAR INTERVALO** [^8].

[^8]: Truncation of posterior predictive t-distribution was done with r package *truncdist* v. 1.0-2.

@@

Although analytic manipulation becomes a bit more messy than obtaining the MLE from the classical approach, it should be obvious by know the benefit of advocating for the later methodology. We gain a coherent inference substructure to include unobserved hypothesis, we get the likelihood resilience insurance for free and in the conjugate case we can trace the interaction between model parameters (which in the mark-down fashion we cannot).

@@

We finally need to adress the parametric space issue as truncation may introduce some distortions of the interest parameter's behavior. To account for a more realistic representation of $\Theta$, in which case we could opt to characterize $X_i$ as $Exp(\theta)$ however the MLE will still have problem **a** and **b** as described in section I.

@@

Given this second conjugate scheme we account for the actual $\Theta$, in fact we gain a substantially simpler math than the normal conjugate case and again we acquire the likelihood resilience insurance for free. Under this new parametrization and setting $\alpha_0 =$ **INSERTAR NÚMERO** and $\beta_0 =$ **INSERTAR NÚMERO** we obtain a 95% credible interval of **INSERTAR INTERVALO** around **INSERTAR ESTIMACIÓN PUNTUAL**.

Again we can directly analyze likelihood resilience to different priors yielding figure 3.

@@

# Wrapping up

Table 1 depicts $\theta$´s different values for in terms of job loss in Mexico with their correspondent regions.

**Table 1**

**Comentar de ls diferencias entre estimadores**

Critical decisions for policy making, such as budget allocation for unemployment insurance or setting target employment levels arise from choosing a particular number from the table above, and any of these translates in a difference of hundreds or maybe millions in terms of cash. Moreover its desirable to have a as robust as possible quantity in which we can rely on.

The frequentist approach, as powerful as it is, may not be the best choice for a setting such the one we are facing now a days[^8]. **CONLUIR**

[^8]: Besides many authors as Mendoza (**INSERTAR AÑO**) have documented many issues with the classical procedures such as failures in the construction of confidence intervals [see](insertar link)
